---
title: "Assignment 1"
author: "Mychala Walker"
format: html
editor: visual
---

## Git and Github

1)  Provide the link to the GitHub repo that you used to practice git from Week 1. It should have:

## Reading Data

2)  Read in the .dta version and store in an object called angell_stata.

```{r}
library(haven)
angell_stata <- read_dta("Assignment1/angell.dta")
```

3)  Read in the .txt version and store it in an object called angell_txt.

```{r}
angell_txt <- read.table(url("https://stats.oarc.ucla.edu/wp-content/uploads/2016/02/angell.txt"), header = FALSE)
```

4)  What are the differences between angell_stata and angell_txt? Are there differences in the classes of the individual columns?

```{r}
identical(angell_stata, angell_txt)
all.equal(angell_stata, angell_txt)

mapply(setdiff, angell_stata, angell_txt)

class(angell_stata)
class(angell_txt)

typeof(angell_stata)
typeof(angell_txt)
```

Using the "typeof" function, it appears that both datasets are identical. However, using the "class" function, the angell_stata dataset was a tbl whereas the angell_txt dataset was a data frame. Also, the header names are different.

4)  Make any updates necessary so that angell_txt is the same as angell_stata.

```{r}

angell_stata <- as.data.frame(angell_stata)
angell_txt <- as.data.frame(angell_txt)

colnames(angell_stata) <- c("city", "morint", "ethhet", "geomob", "region")
colnames(angell_txt) <- c("city", "morint", "ethhet", "geomob", "region")

identical(angell_stata, angell_txt)

```

5)  Describe the Ethnic Heterogeneity variable. Use descriptive statistics such as mean, median, standard deviation, etc. How does it differ by region

```{r}

summary(angell_stata$ethhet)
median(angell_stata$ethhet)
mean(angell_stata$ethhet)
sd(angell_stata$ethhet)
length(angell_stata$ethhet)
```

The ethnic heterogeineity variable is a 43 observation variable. It has a mean of 31.37, a median of 23.70, and a standard deviation of 20.41. It's lower and upper quintiles are 16.9 and 39, respectively. Its lower and upper boundaries are 10.6 and 84.50, respectively.

## Describing Data

## Writing Functions
