---
title: "Assignment3"
author: "Mychala Walker"
format: html
editor: visual
---

```{r, include=FALSE}
library(xml2)
library(rvest)
library(tidyverse)
library(tidytext)
library(dplyr)
library(stringr)
library(wordcloud2) #used to create word cloud
library(devtools) #used to create word cloud
library(tidyr)

```
## Web Scraping

In this assignment, your task is to scrape some information from Wikipedia. We start with the following page about Grand Boulevard, a Chicago Community Area. The ultimate goal is to gather the table “Historical population” and convert it to a data.frame.


```{r, echo=FALSE}
url <-  read_html("https://en.wikipedia.org/wiki/Grand_Boulevard,_Chicago")

table <- url %>% 
  html_elements(css = ".us-census-pop-left") %>% 
  html_table() 

pop <- table[[1]] %>% 
  select(., c(1,2,4)) %>% 
  mutate(across(3, ~ ifelse(row_number() == 1, 0, .))) %>% #change the value in row 1, column 3 to 0
  filter(row_number() != 11) #remove the last row
pop
```

## Expanding to more pages

Can you find the corresponding table in the list of tables that you created earlier? Extract this table as a new
object.

```{r, echo=FALSE}
urls1 <- c("https://en.wikipedia.org/wiki/Grand_Boulevard,_Chicago", 
          "https://en.wikipedia.org/wiki/Armour_Square,_Chicago", 
          "https://en.wikipedia.org/wiki/Douglas,_Chicago",
          "https://en.wikipedia.org/wiki/New_City,_Chicago", 
          "https://en.wikipedia.org/wiki/Washington_Park,_Chicago_(neighborhood)")

# chi <- c("Grand_Boulevard,_Chicago", "Armour_Square,_Chicago", "Douglas,_Chicago", "New_City,_Chicago", "Washington_Park,_Chicago_(neighborhood)")
# 
# links <- paste0("https://en.wikipedia.org/wiki/", chi)


scraper1 <- function(urls) {
  
  all_data <- data.frame()  # Initialize an empty data frame
  
  # Loop through each URL in urls2
  for (url in urls) {
    
    # Read the page content
    page <- read_html(url)
    
    # Extract the table with class .us-census-pop-right
    places <- page %>% 
      html_elements(css = ".us-census-pop-left") %>% 
      html_table(fill = TRUE)  # Convert the HTML table to a list of data frames
    
    # Extract the title of the page
    title <- page %>% 
      html_elements(css = ".mw-page-title-main") %>% 
      html_text()
    
    # Since html_table returns a list, extract the first table from the list
    places_df <- places[[1]]  # Assuming there's only one table
    
    # Add a title column to the places data frame
    places_df$title <- title  # Repeat the title for every row in the table
    
    # Combine this dataset with the full dataset
    all_data <- rbind(all_data, places_df)  # Append the data for this page
  }
  
  return(all_data)
}

place_data1 <- scraper1(urls1) %>% 
  select(., c(1,2,4, 5)) %>% 
  filter(row_number() != c(11, 22, 33, 44, 55))


urls2 <- c("https://en.wikipedia.org/wiki/Oakland,_Chicago", 
          "https://en.wikipedia.org/wiki/Fuller_Park,_Chicago", 
          "https://en.wikipedia.org/wiki/Kenwood,_Chicago", 
          "https://en.wikipedia.org/wiki/Hyde_Park,_Chicago")

# chi2 <- c("Oakland,_Chicago", "Fuller_Park,_Chicago", "Kenwood,_Chicago", "Hyde_Park,_Chicago")
# 
# links2 <- paste0("https://en.wikipedia.org/wiki/", chi2)


scraper2 <- function(urls) {
  
  all_data <- data.frame()  # Initialize an empty data frame
  
  # Loop through each URL in urls2
  for (url in urls) {
    
    # Read the page content
    page <- read_html(url)
    
    # Extract the table with class .us-census-pop-right
    places <- page %>% 
      html_elements(css = ".us-census-pop-right") %>% 
      html_table(fill = TRUE)  # Convert the HTML table to a list of data frames
    
    # Extract the title of the page
    title <- page %>% 
      html_elements(css = ".mw-page-title-main") %>% 
      html_text()
    
    # Since html_table returns a list, extract the first table from the list
    places_df <- places[[1]]  # Assuming there's only one table
    
    # Add a title column to the places data frame
    places_df$title <- title  # Repeat the title for every row in the table
    
    # Combine this dataset with the full dataset
    all_data <- rbind(all_data, places_df)  # Append the data for this page
  }
  
  return(all_data)
}

place_data2 <- scraper2(urls2) %>% 
  select(., c(1,2,4, 5)) %>% 
  filter(!row_number() %in% c(13, 24, 35, 46))

all_places <- rbind(place_data1, place_data2)

all_places
```

## Scraping and Analyzing Text Data

Suppose we wanted to take the actual text from the Wikipedia pages instead of just the
information in the table. Our goal in this section is to extract the text from the body of the
pages, then do some basic text cleaning and analysis.

```{r, echo = FALSE}
# url <-  read_html("https://en.wikipedia.org/wiki/Grand_Boulevard,_Chicago")
# 
# text <- url %>% 
#   html_elements(css = ".vcard+ p") %>% 
#   html_text()
# 
# text

# Define the URLs
urls_full <- c("https://en.wikipedia.org/wiki/Grand_Boulevard,_Chicago",
               "https://en.wikipedia.org/wiki/Oakland,_Chicago", 
               "https://en.wikipedia.org/wiki/Fuller_Park,_Chicago", 
               "https://en.wikipedia.org/wiki/Kenwood,_Chicago", 
               "https://en.wikipedia.org/wiki/Hyde_Park,_Chicago", 
               "https://en.wikipedia.org/wiki/Armour_Square,_Chicago", 
               "https://en.wikipedia.org/wiki/Douglas,_Chicago",
               "https://en.wikipedia.org/wiki/New_City,_Chicago", 
               "https://en.wikipedia.org/wiki/Washington_Park,_Chicago_(neighborhood)")

# Scraper function to extract text and title
scraper3 <- function(urls) {
  
  # Initialize an empty data frame with column names Title and Text
  all_text <- data.frame()
  
  # Loop over each URL in the list
  for (url in urls) {
    
    # Read the HTML content of the page
    page <- read_html(url)
    
    # Extract the text from the paragraph after .vcard class
    text <- page %>% 
      html_elements(css = ".vcard+ p") %>% 
      html_text()
    
    # Extract the title of the page
    title <- page %>% 
      html_elements(css = ".mw-page-title-main") %>% 
      html_text()
    
    # Combine title and text into a data frame and append it to all_text
    all_text <- rbind(all_text, data.frame(Title = title, Text = text, stringsAsFactors = FALSE))
  }
  
  return(all_text)
}

# Call the scraper function and store the extracted data
text_data <- scraper3(urls_full)

# Check the result
text_data
```

Plot the most common words within each location. What are some of the similarities between the locations? What are some of the differences?


```{r, echo = FALSE}
scraper4 <- function(urls) {
  
  # Initialize an empty data frame with column names Title and Text
  all_text <- data.frame()
  
  # Loop over each URL in the list
  for (url in urls) {
    
    # Read the HTML content of the page
    page <- read_html(url)
    
    # Extract the text from the paragraph after .vcard class
    text <- page %>% 
      html_elements(css = ".vcard+ p") %>% 
      html_text()
    
    all_text <- rbind(all_text, 
                      data.frame(Text = text, 
                                 stringsAsFactors = FALSE))
    
  }
  
  return(all_text)
}

text_data2 <- scraper4(urls_full)

# Assuming text_data2 has a column 'Text'
text_data_clean <- text_data2 %>% 
  unnest_tokens(word, Text) %>% #tokenize the data (separate paragraghs and sentences 
  anti_join(stop_words) %>% 
  filter(!grepl("[0-9]+", word))
  
  
text_count <- text_data_clean%>%
  count(word, sort = TRUE) #count the amount of times that each word has been 
                           # used minus removed words
wordcloud <- wordcloud2(text_count)

head(text_count, 7)
wordcloud
```



